---
title: "JM Package Example"
author: "Jared Strauch"
date: "11/14/2022"
output: html_document
---

Preamble

- Package fits joint models for longitudinal and time-to-event data using maximum likelihood

- Applicable in 2 settings:
  1) when focus is on the survival outcome and we wish to account for the effect of an endogenous (aka internal) time-dependent covariates measured with error
  2) when focus is on the longitudinal outcome and we wish to correct for nonrandom dropout

- Basic joint-model-fitting function is jointModel(). 
-- Accepts as main arguments a linear mixed model fitted by function lme() from the nlme package and a Cox model fitted using function coxph() from the survival package



jointModel:
This function fits shared parameter models for the joint modelling of normal longitudinal responses and time-to-event data under a maximum likelihood approach.
```{r}
library(JM)
library(MASS)
library(nlme)
library(splines)
library(survival)
data(aids)
data(aids.id)
# linear mixed model fit with random intercept
# need to include interaction between time and covariate of interest in the mixed model
fitLME <- lme(sqrt(CD4) ~ obstime * drug - drug, random = ~ 1 | patient, data = aids)
# cox model fit
fitCOX <- coxph(Surv(Time, death) ~ drug, data = aids.id, x = TRUE)
# joint model fit with a spline-approximated baseline hazard function
fitJOINT <- jointModel(fitLME, fitCOX, timeVar = "obstime", method = "spline-PH-aGH")
fitJOINT
summary(fitJOINT)
cox <- coxph(Surv(Time, death) ~ drug, data = aids)

loglik <- c(fitJOINT$logLik, cox$loglik[2])
aic <- c(AIC(fitJOINT), AIC(cox))
bic <- c(BIC(fitJOINT),BIC(cox))

fit.check <- data.frame(loglik, aic, bic)
rownames(fit.check) <- c('Joint Model', 'Cox Model')
colnames(fit.check) <- c('Log Likelihood', 'AIC', 'BIC')
fit.check
```
Options for method:
"weibull-AFT-GH" a time-dependent Weibull model under the accelerated failure time formulation is assumed. 

"weibull-PH-GH" a time-dependent relative risk model is postulated with a Weibull baseline risk function. 

"piecewise-PH-GH" a time-dependent relative risk model is postulated with a piecewise constant baseline risk function. 

"spline-PH-GH" a time-dependent relative risk model is assumed in which the log baseline risk
function is approximated using B-splines. If we want to stratify on a covariate we need to use this method. 

"ch-Laplace" an additive model on the log cumulative hazard scale is assumed. 

"Cox-PH-GH" a time-dependent relative risk model is assumed where the baseline risk
function is left unspecified


aucJM:
Using the available longitudinal information up to a starting time point, this function computes an estimate of the prediction error of survival at a horizon time point based on joint models.
```{r}
###Takes forever to run or just doesn't work

# we construct the composite event indicator (transplantation or death)
pbc2$status2 <- as.numeric(pbc2$status != "alive")
pbc2.id$status2 <- as.numeric(pbc2.id$status != "alive")
# we fit the joint model using splines for the subject-specific
# longitudinal trajectories and a spline-approximated baseline
# risk function
lmeFit <- lme(log(serBilir) ~ ns(year, 3), random = list(id = pdDiag(form = ~ ns(year, 3))), data = pbc2)
survFit <- coxph(Surv(years, status2) ~ drug, data = pbc2.id, x = TRUE)
jointFit <- jointModel(lmeFit, survFit, timeVar = "year", method = "piecewise-PH-aGH")
# AUC using data up to year 5 with horizon at year 8
aucJM(jointFit, pbc2, Tstart = 5, Thoriz = 8)
```


Plot:
Produces a variety of plots for fitted joint models.
```{r}
plot(fitJOINT)
```
The plots of the baseline hazard and the cumulative baseline hazard are only produced when the
joint model has been fitted using method = "Cox-PH-GH".

plot.survfitJM:
Produces plots of conditional probabilities of survival.
```{r}
ND <- aids[aids$patient == "141", ]
ss <- survfitJM(fitJOINT, newdata = ND, idVar = "patient", M = 50)
plot(ss)
plot(ss, include.y = TRUE, add.last.time.axis.tick = TRUE, legend = TRUE)
```


Predict:
Calculates predicted values for the longitudinal part of a joint model.
```{r}
# linear mixed model fit
fitLME <- lme(log(serBilir) ~ drug * year, random = ~ year | id, data = pbc2)
# survival regression fit
fitSURV <- survreg(Surv(years, status2) ~ drug, data = pbc2.id, x = TRUE)
# joint model fit, under the (default) Weibull model
fitJOINT <- jointModel(fitLME, fitSURV, timeVar = "year")
# create data frame with different values for drug and year to use in prediction
DF <- with(pbc2, expand.grid(drug = levels(drug),
year = seq(min(year), max(year), len = 100)))

Ps <- predict(fitJOINT, DF, interval = "confidence", return = TRUE)
require(lattice)
xyplot(pred + low + upp ~ year | drug, data = Ps,
type = "l", col = c(2,1,1), lty = c(1,2,2), lwd = 2,
ylab = "Average log serum Bilirubin")
```
When type = "Marginal", this function computes predicted values for the fixed-effects part of
the longitudinal submodel.

When type = "Subject", this functions computes subject-specific predictions for the longitudinal
outcome based on the joint model.


rocJM:
It computes sensitivity, specificity, ROC and AUC measures for joint models.
```{r}
###Takes forever to run or just doesn't work

fitLME <- lme(sqrt(CD4) ~ obstime + obstime:(drug + AZT + prevOI + gender), 
random = ~ obstime | patient, data = aids)
fitSURV <- coxph(Surv(Time, death) ~ drug + AZT + prevOI + gender, data = aids.id, x = TRUE)
fit.aids <- jointModel(fitLME, fitSURV, timeVar = "obstime",
method = "piecewise-PH-aGH")
ND <- aids[aids$patient == "7", ]
roc <- rocJM(fit.aids, dt = c(2, 4, 8), ND, idVar = "patient")
plot(roc, lwd = 2, legend = TRUE)
plot(roc, type = "AUC")
```

dt = a numeric vector indicating the lengths of the time intervals of primary interest within which we want to distinguish between subjects who died within the
intervals from subjects who survived longer than that.


residuals:
Calculates residuals for joint models
```{r}
# residuals for the longitudinal outcome
head(cbind(
"Marginal" = residuals(fitJOINT),
"std-Marginal" = residuals(fitJOINT, type = "stand-Marginal"),
"Subject" = residuals(fitJOINT, type = "Subject"),
"std-Subject" = residuals(fitJOINT, type = "stand-Subject")
))
# residuals for the survival outcome
head(cbind(
"Martingale" = residuals(fitJOINT, process = "Event", type = "Martingale"),
"CoxSnell" = residuals(fitJOINT, process = "Event", type = "CoxSnell")
))
```
if type = "Marginal" these are eij = yij − xij^Tβ, whereas for type = "Subject",
eij = yij − xij^Tβˆ − zij^Tbi, where i denotes the subject and j the measurement, yij the longitudinal responses, xij^T and zij^T the corresponding rows of the fixed and random effects design matrices, respectively, and β and bi denote the fixed effects and random effects components.


Martingale residuals are available for all options for the survival submodel.

Cox-Snell residuals are available for the Weibull model and the additive log cumulative hazard model. 

AFT residuals are only available for the Weibull model.


# Julia's code - same as above, not working

```{r}
if (FALSE) {
fitLME <- lme(sqrt(CD4) ~ obstime * (drug + AZT + prevOI + gender), 
    random = ~ obstime | patient, data = aids)
fitSURV <- coxph(Surv(Time, death) ~ drug + AZT + prevOI + gender, 
    data = aids.id, x = TRUE)
fit.aids <- jointModel(fitLME, fitSURV, timeVar = "obstime", 
    method = "piecewise-PH-aGH")

# the following will take some time to execute...
ND <- aids[aids$patient == "7", ]
roc <- rocJM(fit.aids, dt = c(2, 4, 8), ND, idVar = "patient")
roc
}
```

```{r}
if (FALSE) {
fitLME <- lme(sqrt(CD4) ~ obstime + obstime:(drug + AZT + prevOI + gender), 
    random = ~ obstime | patient, data = aids)
fitSURV <- coxph(Surv(Time, death) ~ drug + AZT + prevOI + gender, 
    data = aids.id, x = TRUE)
fit.aids <- jointModel(fitLME, fitSURV, timeVar = "obstime", 
    method = "piecewise-PH-aGH")

ND <- aids[aids$patient == "7", ]
roc <- rocJM(fit.aids, dt = c(2, 4, 8), ND, idVar = "patient")
plot(roc, lwd = 2, legend = TRUE)
plot(roc, type = "AUC")
}
```

